Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2021-10-18 Manuscript/SS_GAM.tex
Revise the rest of paper

\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}

@@

\abstract{There are proposals that extend the classical generalized
additive models (GAMs) to accommodate the high-dimensional (\(p>>n\))
problem using group penalties and priors. These models rarely smooth the
nonlinear effect properly as the sparse regularization leads to excess
shrinkage on coefficients and inaccurate prediction. Moreover, most of
these GAMs consider an ``all-in-all-out'' approach for function
selection, rendering them difficult to answer if nonlinear effects are
necessary. While some Bayesian models can address these shortcomings,
using Markov chain Monte Carlo (MCMC) algorithms for model fitting
creates a new challenge, scalability. Hence, we propose Bayesian
hierarchical generalized additive models as a solution: we propose a
novel spike-and-slab spline prior for smoothing functions, where a
reparameterization step facilitates smoothing penalty and separate
selection of linear and nonlinear spaces. Two scalable and deterministic
algorithms, EM-Coordinate Descent and EM-Iterative Weighted Least
Squares, are developed to fit the proposed model. Simulation studies and
metabolomics data analyses demonstrate improved performance against
state-of-the-art models, mgcv, COSSO and sparse Bayesian GAM. The
software implementation of the proposed models is freely available via
an R package BHAM.}

@@

Many modern biomedical research, e.g.~sequencing data analysis, electric
health record data analysis, require special treatment of
high-dimensionality, commonly known as \(p >> n\) problem. There is
extensive literature on high-dimensional linear models via penalized
models or Bayesian hierarchical models, see Mallick and Yi
\citep{Mallick2013} for review. These models are built upon a
restrictive and unrealistic assumption, linearity. In classical
statistical modeling, many strategies and models are proposed to relax
the linearity assumption with various degrees of complexity. For
example, variable categorization is a simple and common practice in
epidemiology, but suffers from power and interpretation issues. More
complex models to address nonlinear effects include random forest and
other so-called ``black box'' models \citep{Breiman2001}. These models
are not particularly attractive in biomedical studies, as they lack the
transparency to understand the etiology of diseases. In addition, how to
generalize these models to the high-dimensional setting remains unclear.

Nonparametric regression models are appropriate alternatives to the
``black-box'' models thanks to their balance between model flexibility
and interpretability. Among those, generalized additive models (GAMs),
proposed in the seminal work of Hastie and Tibshirani
\citep{Hastie1987}, grew to be one of the most popular modeling tools.
In a GAM, the response variable, \(Y\), which is assumed to follow some
exponential family distribution, denoted by \(EF(\mu, \phi)\), with mean
\(\mu\) and dispersion \(\phi\), can be modeled with the summation of
smoothing functions, \(B_j(\cdot), j = 1, \dots, p\), of a given
\(p\)-dimensional vector of covariates \(\boldsymbol{x}\), written as \[
 E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)),
\] where \(g^{-1}(\cdot)\) is the inverse of a monotonic link function.
The smoothing functions can take many forms and are estimated using a
pseudo-weighted version of the backfitting algorithm
\citep{Breiman1985}. Nevertheless, the classical GAMs cannot fulfill the
increasing analytic demands for high-dimensional data analysis in
biomedical studies.

There exists some proposals to generalize the classical GAM to
accommodate high-dimensional applications. The regularized models,
branching out from group regularized linear models, are used to fit GAMs
by accounting for the structure introduced when expanding smoothing
functions. Ravikumar et al. \citep{Ravikumar2009} extended the grouped
lasso \citep{Yuan2006} to additive models (AMs); Huang et al.
\citep{Huang2010} further developed adaptive grouped lasso for additive
models; Wang et al. \citep{Wang2007} and Xue \citep{Xue2009}
respectively applied grouped SCAD penalty \citep{Fan2001} to additive
models. Recent Bayesian hierarchical models are also used in the context
of high-dimensional additive models. Various group spike-and-slab priors
combining with computationally intensive Markov chain Monte Carlo (MCMC)
algorithms \citep{Xu2015, Yang2020} are proposed, where the application
on AMs are by-products. Bai et al. \citep{Bai2020} was the first to
apply group spike-and-slab lasso prior to Gaussian AMs using a fast
optimization algorithm, and further generalized the framework to
GAMs\citep{Bai2021}. As these methods focus on addressing the sparsity,
it is very likely to overly penalize the basis function coefficients and
produce inaccurate predictions and curve interpolation, particularly
when complex signals are assumed and large number of knots are used.
\citep{Scheipl2013} In addition, these methods adapt an `all-in-all-out'
strategy, i.e.~either including or excluding the variable completely,
rendering no space for bi-level selection. Scheipl et al.
\citep{Scheipl2012} proposed a spike-and-slab structure prior,
normal-mixture-of-inverse gamma prior, that address the previous
challenges. But the model fitting relies on computational intensive MCMC
algorithmic and creates scalability concern. It would be of special
interest to develop a fast, flexible and accurate generalized additive
model framework.

To address these challenges, we propose a novel Bayesian hierarchical
generalized additive model (BHAM) for high dimensional data analysis.
Specifically, we incorporate smoothing penalties in the model via
re-parameterization of the smoothing function to avoid overly shrinking
basis function coefficients. Smoothing penalties are commonly
implemented in the classical GAMs through smoothing regression splines.
They are quadratic norms of the coefficients and allow locally adaptive
penalties on each smoothing function. A smoothing penalty conditioning
on a smoothing parameter \(\lambda_j\) is a function of the integration
of the second derivative of the spline function, expressed
mathematically as \begin{equation}\label{eq:smoothpen}
  \text{pen}\left[B_j(x)\right] = \lambda_j \int B^{\prime\prime}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j ,
\end{equation} where \(\boldsymbol{S}_j\) is a known smoothing penalty
matrix, and \(\boldsymbol{\beta}_j\) are the basis function
coefficients. Smoothing penalties were also previously used in the
spike-and-slab GAM \citep{Scheipl2012} and the sparsity-smoothness
penalty \citep{Meier2009}. Moreover, incorporating the smoothing penalty
allows the separation of the linear space of smoothing functions from
the nonlinear space. We then impose a new bi-part spike-and-slab spline
prior on the smoothing functions for bi-level selection such that the
linear and nonlinear spaces of smoothing functions can be selected
separately. The prior setup encourages a flexible solution, rending one
of three possibilities: no effect, only linear effect, or nonlinear
effect of a predictor. In addition, two scalable optimization-based
algorithms are developed for high and
ultrahigh-dimensional\citep{Fan2009} settings respectively.

The proposed framework, BHAM, differs from previous spike-and-slab based
GAMs, i.e.~the spike-and-slab GAM \citep{Scheipl2012} and the SB-GAM
\citep{Bai2021} in many ways. First of all, the proposed spike-and-slab
spline prior is a spike-and-slab lasso type prior using independent
mixture double exponential distribution, compared to spike-and-slab GAM.
Spike-and-slab lasso priors provide computational convenience during
model fitting by using optimization algorithms instead of computational
intensive sampling algorithms. They make fitting high- and
ultrahigh-dimensional models more feasible without sacrificing
performance in prediction and variable selection. Secondly, the
optimization algorithms introduced in this article include EM-Coordinate
Descent (EM-CD) algorithm and EM-Iterative weighted least square
(EM-IWLS) algorithm for different utilities. The applications of the two
aforementioned algorithms are fruitful in Bayesian hierarchical models
for high-dimensional data. \citep{Yi2012, Rockova2014a, Rockova2018}
Even though SB-GAM with a group spike-and-slab lasso prior also uses
EM-CD for model fitting, the proposed EM-IWLS provides uncertainty
measures that could be further used for confidence bound calculation and
hypothesis testing. EM-IWLS complements the lack of inference potential
from EM-CD in a median and high dimensional setting. Furthermore, the
proposed model addresses the incapability of bi-level selection in
SB-GAM. Lastly, the R implementation of the Bayesian hierarchical
additive model is available publicly via BHAM at
\url{https://github.com/boyiguo1/BHAM}, making transnational science
more accessible.

In Section \ref{sec:BHAM}, we establish the Bayesian hierarchical
generalized additive model, introduce the proposed spike-and-slab spline
priors, and describe the two fast-fitting algorithms. In Section
\ref{sec:sim}, we compare the proposed framework to state-of-art models,
mgcv, COSSO and sparse Bayesian GAM via Monte Carlo simulation studies.
Analyses of two metabolomics datasets are presented in Section
\ref{sec:real_data}. Conclusion and discussions are given in Section
\ref{sec:concl}.

@@

Following the GLM notation introduced in Section \ref{sec:intro}, we
have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation} and the data distribution is \begin{equation}
f(\boldsymbol{Y} = \boldsymbol{y}| \boldsymbol{\beta}, \phi) = \prod\limits^n_{i=1}f( Y = y_i|\boldsymbol{\beta}, \phi),\nonumber
\end{equation} with smoothing functions \(B_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The basis function matrix, i.e.~the design
matrix derived from the smoothing function \(B_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+3\)
elements column vector. Its corresponding smoothing penalty matrix is
denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\) denotes the
basis function coefficients for the \(j\)th variable such that
\(B_j(x_j) = \boldsymbol{\beta}_j^T \boldsymbol{X}_j\). With slight
abuse of notation, we denote vectors and matrices in bold fonts
\(\boldsymbol{\beta}, \boldsymbol{X}\) with conformable dimensions,
where scalar and random variables are denoted in with unbold fonts
\(\beta, X\). The matrix transposing operation is denoted with a
superscript \(^T\). To note, the proposed model can include parametric
forms of variables in the model, treating the parametric function as a
special case of the smoothing function, e.g.~\(B_j(x_j) = x_j\) with the
smoothing penalty matrix defined as
\(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

To encourage proper smoothing of the functions, we adopt the idea of
smoothing penalties from smoothing spline models. The basic idea is to
set up a smoothing penalty, described in Equation (\ref{eq:smoothpen}),
in the prior density function. However, the direct integration of
smoothing penalty with sparsity penalty is not obvious. Marra and Wood
\citep{Marra2011} proposed a re-parameterization procedure to
accommodate the smoothing penalty implicitly. Given the smoothing
penalty matrix \(\boldsymbol{S}_j\) is symmetric and positive
semi-definite for univariate smoothing functions, we apply
eigendecomposition on the penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\) ,
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
elements of zeros on the diagonal, where the zeros are associated with
the linear space of the smoothing function. For the most popular
smoothing function, cubic splines, the dimension of the linear space is
one. Hereafter, we focus on discussing a uni-dimensional linear space
for simplicity; however, it generalizes easily to the cases where the
linear space is multidimensional. We further write the orthonormal
matrix
\(\boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{U}^0 : \boldsymbol{U}^{*}\end{bmatrix}\)
containing the eigenvectors as columns in the corresponding order to
\(\boldsymbol{D}\). That is, \(\boldsymbol{U}\) contains the
eigenvectors \(U^0\) with zero eigenvalues for the linear space and
\(\boldsymbol{U}^{*}\) contains the eigenvectors (as columns) for the
non-zero eigenvalues, i.e.~the non-linear space. We multiply the basis
function matrix \(\boldsymbol{X}\) with the orthonormal matrix
\(\boldsymbol{U}\) for the new design matrix
\({\boldsymbol{X}}^\text{repa}= \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} X^0 : \boldsymbol{X}^{*} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{*}\) by the
non-zero eigenvalues of \(\boldsymbol{D}\) such that the new basis
function matrix \(\boldsymbol{X}^\ast\) can receive uniform penalty on
each of its dimensions. With slight abuse of the notation, we drop the
superscript \(^\text{repa}\) and denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} X_j^0 : \boldsymbol{X}_j^{*} \end{bmatrix}\)
as the basis function matrix for the \(j\)th variable after the
re-parameterization. A spline function can be expressed in the matrix
form \[
B_j(x_j) = B_j^0(x_j) + B_j^*(x_j) = \beta_j X^0_j + \boldsymbol{\beta_j^*}^T \boldsymbol{X}_j^*,
\] and the generalized additive model in Equation (\ref{eq:gam}) be
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[\beta_0 + \sum\limits^p_{j=1} (\beta_j X^0_j + {\boldsymbol{\beta}_j^*}^T \boldsymbol{X}_j^*)\right],
\end{equation} where the coefficients
\(\boldsymbol{\beta}_j \equiv \begin{bmatrix} \beta_j : \boldsymbol{\beta}^*_j \end{bmatrix}\)
is an augmentation of the coefficient scalar \(\beta_j\) of linear space
and the coefficient vector \(\boldsymbol{\beta}^*_j\) of non-linear
space.

The re-parameterization step sets up the foundation of the proposed
model and provides three-fold benefits. First of all, the
re-parameterization integrates the smoothing penalty into the design
matrix, and encourages models to properly smooth the nonlinear function
in addition to the sparse penalty for function selection. Secondly, the
eigendecomposition of the smoothing penalty allows the isolation of the
linear from the nonlinear space, improving the feasibility of bi-level
function selection and inference. The eigendecomposition facilitates the
construction of orthonormal design matrix, which makes imposing
independent priors on the coefficients possible. This reduces the
computational complexity compared to using a multivariate priors, and
greatly broadens the choices of priors and further model choices. Last
but not least, the scaling step of \(\boldsymbol{X}^*\) further
simplifies the choice of priors, from column-specific priors to a
unified prior.

\hypertarget{spike-and-slab-smooth-priors}{%
\subsection{Spike-and-Slab Smooth
Priors}\label{spike-and-slab-smooth-priors}}

The family of spike-and-slab (SS) priors regression models dominates
Bayesian high-dimensional analysis for its utility in outcome prediction
and variable selection. We defer to Bai et al. \citep{Bai2021Review} for
an in-depth introduction to spike-and-slab priors. To summarize,
spike-and-slab priors are a family of mixture distributions that
comprises a skinny spike density \(f_{\text{spike}}(\cdot)\) for weak
signals and a flat slab density \(f_{\text{slab}}(\cdot)\) for strong
signals, mathematically \[
 \beta|\gamma \sim (1-\gamma)f_{\text{spike}}(\beta) + \gamma f_{\text{slab}}(\beta).
\] The most distinct feature of SS priors is that it is conditioned on a
latent binary variable \(\gamma \in \{0,1\}\) that indicates whether the
variable \(x\) is included in the model. There are various
spike-and-slab priors depending on the choice for the spike density
\(f_{\text{spike}}(\cdot)\) and the slab density
\(f_{\text{slab}}(\cdot)\), see George and McCulloch
\citep{George1993, George1997}; Chipman \citep{Chipman1996} for grouped
variables; Brown et al. \citep{Brown1998} for multivariate outcomes;
Ishwaran and Rao \citep{Ishwaran2005}; Clyde and George
\citep{Clyde2004} and reference therein.

>>>>

\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}

@@

\abstract{There are proposals that extend the classical generalized
additive models (GAMs) to accommodate high-dimensional data (\(p>>n\))
using group sparse regularization. However, the sparse reuglarization
may induce excess shrinkage when estimating smoothing functions,
damaging predictive performance. Moreover, most of these GAMs consider
an ``all-in-all-out'' approach for functional selection, rendering them
difficult to answer if nonlinear effects are necessary. While some
Bayesian models can address these shortcomings, using Markov chain Monte
Carlo algorithms for model fitting creates a new challenge, scalability.
Hence, we propose Bayesian hierarchical generalized additive models as a
solution: we consider the smoothing penalty for proper shrinkage of
curve interpolation and separation of smoothing function linear and
nonlinear spaces. A novel spike-and-slab spline prior is proposed to
select components of smoothing functions. Two scalable and deterministic
algorithms, EM-Coordinate Descent and EM-Iterative Weighted Least
Squares, are developed for different utilities. Simulation studies and
metabolomics data analyses demonstrate improved predictive or
computational performance against state-of-the-art models, mgcv, COSSO
and sparse Bayesian GAM. The software implementation of the proposed
models is freely available via an R package BHAM.}

@@

Many modern biomedical research, e.g.~sequencing data analysis, electric
health record data analysis, require special treatment of
high-dimensionality, commonly known as \(p >> n\) problem. There is
extensive literature on high-dimensional linear models via penalized
models or Bayesian hierarchical models, see Mallick and Yi
\citep{Mallick2013} for review. These models are built upon a
restrictive and unrealistic assumption, linearity. In classical
statistical modeling, many strategies and models are proposed to relax
the linearity assumption with various degrees of complexity. For
example, variable categorization is a simple and common practice in
epidemiology, but suffers from power and interpretation issues. More
complex models to address nonlinear effects include random forest and
other so-called ``black box'' models \citep{Breiman2001}. These models
are useful for statistical prediction but do not estimate parameters
relevant to the data generation process that one can draw inferences
from. In addition, how to generalize these models to the
high-dimensional setting remains unclear.

Nonparametric regression models are appropriate alternatives to the
``black-box'' models thanks to their balance between model flexibility
and interpretability. Among those, generalized additive models (GAMs),
proposed in the seminal work of Hastie and Tibshirani
\citep{Hastie1987}, grew to be one of the most popular modeling tools.
In a GAM, the response variable, \(Y\), which is assumed to follow some
exponential family distribution with mean \(\mu\) and dispersion
\(\phi\), can be modeled with the summation of smoothing functions,
\(B_j(\cdot), j = 1, \dots, p\), of a given \(p\)-dimensional vector of
covariates \(\boldsymbol{x}\), written as \[
 E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)),
\] where \(g^{-1}(\cdot)\) is the inverse of a monotonic link function.
The smoothing functions can take many forms and are estimated using a
pseudo-weighted version of the backfitting algorithm
\citep{Breiman1985}. Nevertheless, the classical GAMs cannot fulfill the
increasing analytic demands for high-dimensional data analysis in
biomedical studies.

There exists some proposals to generalize the classical GAM to
accommodate high-dimensional applications. The regularized models,
branching out from group regularized linear models, are used to fit GAMs
by accounting for the structure introduced when expanding smoothing
functions. Ravikumar et al. \citep{Ravikumar2009} extended the grouped
lasso \citep{Yuan2006} to additive models (AMs); Huang et al.
\citep{Huang2010} further developed adaptive grouped lasso for additive
models; Wang et al. \citep{Wang2007} and Xue \citep{Xue2009}
respectively applied grouped SCAD penalty \citep{Fan2001} to additive
models. Recently Bayesian hierarchical models are also used in the
context of high-dimensional additive models. Various group
spike-and-slab priors combining with computationally intensive Markov
chain Monte Carlo (MCMC) algorithms \citep{Xu2015, Yang2020} are
proposed, where the application on AMs are by-products. Bai et al.
\citep{Bai2020} was the first to apply group spike-and-slab lasso prior
to Gaussian AMs using a fast optimization algorithm, and further
generalized the framework to GAMs\citep{Bai2021}. Focus on addressing
the sparsity, these methods can overly penalize the basis function
coefficients and produce inaccurate predictions and curve interpolation,
particularly when complex signals are assumed and large number of knots
are used. \citep{Scheipl2013} In addition, these methods adapt an
`all-in-all-out' strategy, i.e.~either including or excluding the
variable completely, rendering no space for bi-level selection. Scheipl
et al. \citep{Scheipl2012} proposed a spike-and-slab structure prior
that address the previous challenges. But the model fitting relies on
computational intensive MCMC algorithms and creates scalability concern.
It would be of special interest to develop a fast, flexible and accurate
generalized additive model framework.

To address these challenges, we propose a novel Bayesian hierarchical
generalized additive model (BHAM) for high dimensional data analysis.
Specifically, we incorporate smoothing penalties in the model via
re-parameterization of the smoothing function to avoid overly shrinking
basis function coefficients. Smoothing penalties are commonly
implemented in the classical GAMs through smoothing regression splines.
They are quadratic norms of the coefficients and allow locally adaptive
penalties on each smoothing function. A smoothing penalty conditioning
on a smoothing parameter \(\lambda_j\) is a function of the integration
of the second derivative of the spline function, expressed
mathematically as \begin{equation}\label{eq:smoothpen}
  \text{pen}\left[B_j(x)\right] = \lambda_j \int B^{\prime\prime}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j ,
\end{equation} where \(\boldsymbol{S}_j\) is a known smoothing penalty
matrix, and \(\boldsymbol{\beta}_j\) are the basis function
coefficients. Smoothing penalties were also previously used in the
spike-and-slab GAM \citep{Scheipl2012} and the sparsity-smoothness
penalty \citep{Meier2009}. Moreover, incorporating the smoothing penalty
allows the separation of the linear space of a smoothing function from
the nonlinear space. We then impose a new two-part spike-and-slab spline
prior on the smoothing functions for bi-level selection such that the
linear and nonlinear spaces of smoothing functions can be selected
separately. The prior setup encourages a flexible solution, rendering
one of three possibilities for each predictor: no effect, only linear
effect, or linear and nonlinear effects. In addition, two scalable
optimization-based algorithms, EM-Coordinate Descent (EM-CD) algorithm
and EM-Iterative weighted least square (EM-IWLS) algorithm, are
developed and implemented in an publicly available R package
\texttt{BHAM} via \url{https://github.com/boyiguo1/BHAM}, making
translational science more accessible.

The proposed framework, BHAM, differs from previous spike-and-slab based
GAMs, i.e.~the spike-and-slab GAM \citep{Scheipl2012} and the SB-GAM
\citep{Bai2021} in three ways. First of all, the proposed spike-and-slab
spline prior is a spike-and-slab lasso type prior using independent
mixture double exponential distribution, compared to spike-and-slab GAM
that uses normal-mixture-of-inverse gamma prior. Spike-and-slab lasso
priors provide computational convenience during model fitting by using
optimization algorithms instead of intensive sampling algorithms. They
make fitting high-dimensional models more feasible without sacrificing
performance in prediction and variable selection. Secondly, SB-GAM uses
a group spike-and-slab lasso prior with an EM-CD algorithm to fit the
model. While both methods use the combination of expectation
maximization algorithm and coordient descent algorithm, there are subtle
difference in the implementation due to the difference in prior
specification. The proposed model sets up independent priors among basis
function coefficients after the re-parameterization step, which provides
some advantage in computation. In addition, we offer an alternative
algorithm, EM-IWLS, that can provide variance-covariance matrix of the
coefficients. Last but not least, the proposed model addresses the
incapability of bi-level selection in SB-GAM.

In Section \ref{sec:BHAM}, we establish the Bayesian hierarchical
generalized additive model, introduce the proposed spike-and-slab spline
priors, and describe the two fast-fitting algorithms. In Section
\ref{sec:sim}, we compare the proposed framework to state-of-the-art
models, mgcv, COSSO and sparse Bayesian GAM via Monte Carlo simulation
studies. Analyses of two metabolomics datasets are presented in Section
\ref{sec:real_data}. Conclusion and discussions are given in Section
\ref{sec:concl}.

@@

Following the GLM notation introduced in Section \ref{sec:intro}, we
have a generalized additive model with link function \(g(\cdot)\) and
linear predictor \begin{equation}\label{eq:gam}
\eta = \beta_0 + \sum\limits^p_{j=1}B_j(x_j) = \beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j,
\end{equation} with smoothing functions \(B_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The outcome \(Y\) follows a exponential family
distribution with density function \(f(y)\), mean \(\mu = g^{-1}(\eta)\)
and dispersion parameter \(\phi\), and the data distribution is
\begin{equation}
f(\boldsymbol{Y} = \boldsymbol{y}| \boldsymbol{\beta}, \phi) = \prod\limits^n_{i=1}f( Y = y_i|\boldsymbol{\beta}, \phi),\nonumber
\end{equation} The basis function matrix, i.e.~the design matrix derived
from the smoothing function \(B_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function, and is
denoted as \(K_j\) for \(x_j\). Its corresponding smoothing penalty
matrix is denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\)
denotes the basis function coefficients for the \(j\)th variable such
that \(B_j(x_j) = \boldsymbol{\beta}_j^T \boldsymbol{X}_j\). With slight
abuse of notation, we denote vectors and matrices in bold fonts
\(\boldsymbol{\beta}, \boldsymbol{X}\) with conformable dimensions,
where scalar and random variables are denoted in unbold fonts
\(\beta, X\). The matrix transposing operation is denoted with a
superscript \(^T\). To note, the proposed model can include parametric
forms of variables in the model, treating the parametric function as a
special case of the smoothing function, e.g.~\(B_j(x_j) = x_j\) with the
smoothing penalty matrix defined as
\(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

To encourage proper smoothing of additive functions, we adopt the idea
of smoothing penalties from smoothing spline models. The basic idea is
to set up a smoothing penalty, described in Equation
(\ref{eq:smoothpen}), in the prior density function. However, the direct
integration of smoothing penalty with sparsity penalty is not obvious.
Marra and Wood \citep{Marra2011} proposed a re-parameterization
procedure to accommodate the smoothing penalty implicitly. Given the
smoothing penalty matrix \(\boldsymbol{S}_j\) is symmetric and positive
semi-definite for univariate smoothing functions, we apply
eigendecomposition on the penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\) ,
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
elements of zeros on the diagonal, where the zeros are associated with
the linear space of the smoothing function. For the most popular
smoothing function, cubic splines, the dimension of the linear space is
one. Hereafter, we focus on discussing a uni-dimensional linear space
for simplicity; however, it generalizes easily to the cases where the
linear space is multidimensional. We further write the orthonormal
matrix
\(\boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{U}^0 : \boldsymbol{U}^{*}\end{bmatrix}\)
containing the eigenvectors as columns in the corresponding order to
\(\boldsymbol{D}\). That is, \(\boldsymbol{U}\) contains the
eigenvectors \(U^0\) with zero eigenvalues for the linear space and
\(\boldsymbol{U}^{*}\) contains the eigenvectors (as columns) for the
non-zero eigenvalues, i.e.~the non-linear space. We multiply the basis
function matrix \(\boldsymbol{X}\) with the orthonormal matrix
\(\boldsymbol{U}\) for the new design matrix
\({\boldsymbol{X}}^\text{repa}= \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} X^0 : \boldsymbol{X}^{*} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{*}\) by the
non-zero eigenvalues of \(\boldsymbol{D}\) such that the new basis
function matrix \(\boldsymbol{X}^\ast\) can receive uniform penalty on
each of its dimensions. With slight abuse of the notation, we drop the
superscript \(^\text{repa}\) and denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} X_j^0 : \boldsymbol{X}_j^{*} \end{bmatrix}\)
as the basis function matrix for the \(j\)th variable after the
re-parameterization. A spline function can be expressed in the matrix
form \[
B_j(x_j) = B_j^0(x_j) + B_j^*(x_j) = \beta_j X^0_j + \boldsymbol{\beta_j^*}^T \boldsymbol{X}_j^*,
\] and the generalized additive model in Equation (\ref{eq:gam}) now is
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[\beta_0 + \sum\limits^p_{j=1} (\beta_j X^0_j + {\boldsymbol{\beta}_j^*}^T \boldsymbol{X}_j^*)\right],
\end{equation} where the coefficients
\(\boldsymbol{\beta}_j \equiv \begin{bmatrix} \beta_j : \boldsymbol{\beta}^*_j \end{bmatrix}\)
is an augmentation of the coefficient scalar \(\beta_j\) of linear space
and the coefficient vector \(\boldsymbol{\beta}^*_j\) of non-linear
space.

The re-parameterization step sets up the foundation of the proposed
model and provides three benefits. First of all, the re-parameterization
integrates the smoothing penalty into the design matrix, and encourages
models to properly smooth the nonlinear function in addition to the
sparse penalty for functional selection. Secondly, the
eigendecomposition of the smoothing penalty allows the isolation of the
linear space from the nonlinear space, improving the feasibility of
bi-level functional selection. The eigendecomposition facilitates the
construction of orthonormal design matrix, which makes imposing
independent priors on the coefficients possible. This reduces the
computational complexity compared to using a multivariate priors, and
greatly broadens the choices of priors and further model choices. Last
but not least, the scaling step of \(\boldsymbol{X}^*\) further
simplifies the choice of priors, from column-specific priors to a
unified prior.

\hypertarget{spike-and-slab-spline-priors}{%
\subsection{Spike-and-Slab Spline
Priors}\label{spike-and-slab-spline-priors}}

The family of spike-and-slab (SS) regression models is one of most
commonly used models in high-dimensional data analysis for its utility
in outcome prediction and variable selection. We defer to Bai et al.
\citep{Bai2021Review} for an in-depth introduction to spike-and-slab
priors. To summarize, spike-and-slab priors are a family of mixture
distributions that comprises a skinny spike density
\(f_{\text{spike}}(\cdot)\) for weak signals and a flat slab density
\(f_{\text{slab}}(\cdot)\) for strong signals, mathematically \[
 \beta|\gamma \sim (1-\gamma)f_{\text{spike}}(\beta) + \gamma f_{\text{slab}}(\beta).
\] The most distinct feature of SS priors is that it is conditioned on a
latent binary variable \(\gamma \in \{0,1\}\) that indicates whether the
variable \(x\) is included in the model. There are various
spike-and-slab priors depending on the choice for the spike density
\(f_{\text{spike}}(\cdot)\) and the slab density
\(f_{\text{slab}}(\cdot)\), see George and McCulloch
\citep{George1993, George1997}; Chipman \citep{Chipman1996} for grouped
variables; Brown et al. \citep{Brown1998} for multivariate outcomes;
Ishwaran and Rao \citep{Ishwaran2005}; Clyde and George
\citep{Clyde2004} and reference therein. Two of most popular spike and
slab priors are spike-and-slab normal prior \citep{George1993} and
spike-and-slab double exponential prior\citep{Rockova2018}.

The major criticism of early spike-and-slab models is being
computationally prohibitive. \citep{Bai2021Review} Since then, many
studies focus on alleviating the computational burden that sampling
algorithms bear, which include EMVS based on spike-and-slab normal
prior\citep{Rockova2014a} and spike-and-slab Lasso (SSL)
\citep{Rockova2018b, Rockova2018}. Particularly, the development of SSL
model substantially improves the scalability of SS models, setting up
the theoretical foundation for generalized models in -omics data
analysis \citep{Tang2017a, Tang2017, Tang2018, Tang2019}. The SSL prior
is composed of two double exponential distributions with mean 0 and
different dispersion parameters, \(0 < s_0 < s_1\), mathematically,
\begin{equation} 
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber
\end{equation} Given that both double exponential distributions have a
mean of 0 and the latent indicator \(\gamma\) can only take the value of
0 or 1, the mixture double exponential distribution can be formulated as
one single double exponential density, \begin{equation} \label{eq:ssl}
\beta | \gamma \sim DE(0, S), 0 < s_0 < s_1,
\end{equation} with the scale parameter
\(S = (1-\gamma)s_0 + \gamma s_1\). The SSL also mitigates the problem
of EMVS where the weak signals are not shrink to zero, and hence is
preferred in high-dimensional data analysis. We notice that Bai
\citep{Bai2021} is the first to apply spike-and-slab lasso prior in the
GAM framework, where the densities of the spike and slab components take
the group lasso density \citep{Xu2015} and limits to an
``all-in-all-out'' strategy for functional selection.

\hypertarget{two-part-spike-and-slab-lasso-prior}{%
\subsubsection{Two-part Spike-and-Slab Lasso
Prior}\label{two-part-spike-and-slab-lasso-prior}}
